A Mechanised Library for Generic (Co)Programming

### 2024-01-18 paper/paper.tex
Notes to introduction + Related work

\hypersetup{
  linktocpage,
  colorlinks,
  citecolor=BlueViolet,
  filecolor=red,
  linkcolor=Blue,
  urlcolor=BrickRed
}
\newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
\newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
\newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
\newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
\newcommand{\bnfeq}{\mathrel{::=}}
\newcommand{\defeq}{\triangleq}
\newcommand{\rul}[3]{\frac{#2}{#3}\;  {\textrulelabel{#1}}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\jud}[3]{#1 \vdash #2 : #3}
\newcommand{\bden}[1]{\llparenthesis#1 \rrparenthesis}
\newcommand{\bigslant}[2]{{\raisebox{.2em}{$#1$}\left/\raisebox{-.2em}{$#2$}\right.}}
\newcommand{\quotient}[2]{\bigslant{#1}{#2}}

@@

\maketitle
\section{Introduction}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or productivity.
For recursion, this is done by trying to (automatically) infer which argument in
the recursive call gets smaller with respects to the original input argument.
For productivity, the algorithm checks that the (co)recursive call appears
directly under a constructor to make sure that this function always produce at
least one element after each recursive step.

@@

@@

One of the challenges here is to be able to keep the theory of recursions
schemes as general as possible while maintaining its practical usefulness. In
order to do this, we have to make some design choices which differ a bit from
the general theory.  For instance, formalising the conjugate rule would require
us to mechanise a non-trivial fragment of category theory inside the prover's
logic.  Secondly, the general conjugate rule is a statement of correspondence of
arrows rather than a proof that these programs actually exists. In functional
programming languages like for instance Haskell this is not a problem, since
every functor is strong and has a fixed-point.  On the other hand, in a type
theory we have to restrict to those so-called polynomial functors. In this work
we do not make use of these assumptions. By restricting ourselves to polynomial
functors given in terms of containers~\cite{AbbottAG05} we are able to prove the
recursion schemes within the proof assistant and at the same time to maintain a
certain level of data abstraction.

@@

\begin{itemize}
  \item McBride devised a generalisation of Bove and Capretta's general
recursive function monad~\cite{McBride15, BoveC01}
  \item Guarded recursion for productive definitions and general recursion
~\cite{AtkeyM13, PaviottiMB15}
  \item PACO library~\cite{HurNDV13}
  \item Sized Types \cite{AbelPTS13, AbelP16}
\end{itemize}

>>>>

\hypersetup{
  linktocpage,
  colorlinks,
  citecolor=BlueViolet,
  filecolor=red,
  linkcolor=Blue,
  urlcolor=BrickRed
}
\newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
\newcommand{\dcas}[1]{\textcolor{red}{\textsc{David}: #1}}
\newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
\newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
\newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
\newcommand{\bnfeq}{\mathrel{::=}}
\newcommand{\defeq}{\triangleq}
\newcommand{\rul}[3]{\frac{#2}{#3}\;  {\textrulelabel{#1}}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\jud}[3]{#1 \vdash #2 : #3}
\newcommand{\bden}[1]{\llparenthesis#1 \rrparenthesis}
\newcommand{\bigslant}[2]{{\raisebox{.2em}{$#1$}\left/\raisebox{-.2em}{$#2$}\right.}}
\newcommand{\quotient}[2]{\bigslant{#1}{#2}}

@@

\maketitle
\section{Introduction}

Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or productivity.
For recursion, this is done by trying to (automatically) infer which argument in
the recursive call gets smaller with respects to the original input argument.
For productivity, the algorithm checks that the (co)recursive call appears
directly under a constructor to make sure that this function always produce at
least one element after each recursive step.

@@

\dcas{My take would be different here from now on. See below:}

Recursion schemes enjoy generic algebraic laws for program manipulation. Many
common optimisations and transformations are instances of such laws. Examples
of this include shortcut deforestation~\cite{TakanoM95}, and
parallelisations~\cite{Gibbons96:Third}. \dcas{TODO: add more examples?}

With this work, we study the formalisation of hylomorphism laws in proof
assistants like Coq to derive efficient and correct recursive algorithms by
program calculation techniques. Furthermore, our encoding of hylomorphisms
allows Coq's code extraction mechanism to produce OCaml code that resembles
what a programmer would have writen manually.

\dcas{END (TODO: merge with Marco's. I would omit (or take parts of the) below)}

@@

\dcas{Note: the below is useful to keep in the intro}
One of the challenges here is to be able to keep the theory of recursions
schemes as general as possible while maintaining its practical usefulness. In
order to do this, we have to make some design choices which differ a bit from
the general theory.  For instance, formalising the conjugate rule would require
us to mechanise a non-trivial fragment of category theory inside the prover's
logic.  Secondly, the general conjugate rule is a statement of correspondence of
arrows rather than a proof that these programs actually exists. In functional
programming languages like for instance Haskell this is not a problem, since
every functor is strong and has a fixed-point.  On the other hand, in a type
theory we have to restrict to those so-called polynomial functors. In this work
we do not make use of these assumptions. By restricting ourselves to polynomial
functors given in terms of containers~\cite{AbbottAG05} we are able to prove the
recursion schemes within the proof assistant and at the same time to maintain a
certain level of data abstraction.

@@

\begin{itemize}
  \item McBride devised a generalisation of Bove and Capretta's general
recursive function monad~\cite{McBride15, BoveC01}
  \item Guarded recursion for productive definitions and general recursion
~\cite{AtkeyM13, PaviottiMB15}
  \item PACO library~\cite{HurNDV13}
  \item Sized Types \cite{AbelPTS13, AbelP16}
  \item IMPORTANT!!!!! Coq extraction from recursion
    schemes~\cite{larchey2022braga} \dcas{Check bibtex}
  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
  \item Recursion schemes in Coq (but not hylos)~\cite{MurataE19}
  \item EVEN MORE IMPORTANT!!! 2023 POPL paper on divide and conquer code in
    Coq~\cite{AbreuDHJMS23}.  Some differences to point out: they require
    functional-extensionality axiom. We are axiom-free. Little discussion about
    extraction. They focus mostly on erasing termination proofs, but they do
    not discuss "what does their extracted code look like". Their
    representation of Functors leads me to think that they will produce loads
    of "Obj.magic", but we'd need to check this. 
\end{itemize}
